A New Hybrid Machine Learning Method for Stellar Parameter Inference

### 2025-04-08 paper/paper2/paper.tex
Unbold stuff

\documentclass[twocolumn, linenumbers]{aastex631}
\usepackage{graphicx}
\usepackage{physics}
\usepackage[super]{nth}
\usepackage{enumitem}
\usepackage[all]{hypcap}

@@

\begin{abstract}
    The advent of machine learning (ML) is revolutionary to numerous scientific disciplines, with a growing number of examples in astronomical spectroscopic inference, as ML is more powerful than traditional techniques.
    Here we introduce a hybrid ML (HML) method combin\textbf{ing} automatic differentiation, interpolation, and Bayesian optimization to infer stellar parameters \textbf{of} stellar spectra.
    We study $T_{\mathrm{eff}}$, $\log(g)$, and [Fe/H], but this method \textbf{could be extended} to other \textbf{parameters} such as [$\alpha$/Fe] \textbf{(alpha element abundance)}, C/O \textbf{(carbon-oxygen ratio)}, and $f_{sed}$ \textbf{(sedimentation efficiency)}.
    We first use \texttt{blase} \textbf{to} semi-empirical\textbf{ly recast} spectra \textbf{into} sets of \textbf{Voigt profiles}.
    \texttt{blase} is \textbf{run} on 1,314 \textbf{synthetic} spectra from \textbf{a rectilinear subset of} the PHOENIX model grid \textbf{($T_\text{eff}$: [2300, 12000] K, $\log(g)$: [2, 6], [Fe/H]: [-0.5, 0], $\lambda$: [8038, 12849] \AA)}.
    For \textbf{128,723 detected features}, we map stellar parameters to spectral line parameters using linear interpolation.
    \textbf{This} create\textbf{s} the PHOENIX generator, \textbf{enabling parallelized spectral synthesis}.
    Gaussian Process minimization is used to infer stellar parameters by minimizing \textbf{a} root-mean-square loss \textbf{function}.
    Testing \textbf{210 noise-free models ($T_{\mathrm{eff}}$: [3000, 11000] K, $\log(g)$: [2, 6], [Fe/H]: [-0.5, 0]), we find inference errors:} $T_{\mathrm{eff}}$\textbf{: 93 K, $\log(g)$: 0.24, and [Fe/H]: 0.056 for $T_\text{eff} < $ 7000 K, and $T_{\mathrm{eff}}$: 347 K, $\log(g)$: 0.26, and [Fe/H]: 0.16 for $T_\text{eff} \ge$ 7000 K}.
    \textbf{We also upload online} an archive of \texttt{blase} \textbf{models} of the PHOENIX subset.
    This proof-of-concept \textbf{study} shows that semi-empirical HML is a viable alternative to traditional approaches \textbf{in spectroscopic inference}.
\end{abstract}

@@

Modern astronomical spectroscopy \textbf{takes advantage of intuitive, performant spectral models such as KORG \citep{korg}, however it} would \textbf{further} benefit from \textbf{the paradigm of interpretability}.
The determination of fundamental properties, the creation of Extreme Precision Radial Velocity (EPRV) templates, and the application of composite spectral fitting could all leverage such an innovation.
This ambitious aim may be broadly referred to as a ``foundational spectroscopy model for astrophysics'', in reference to the same category of models used for large language models (LLMs) in artificial intelligence (AI).
There are many challenges with creating such a foundational spectral model for astrophysics.
Physical inputs into the spectral modeling process are imperfect\textbf{, and} simplifying assumptions in stellar (and substellar) atmospheres are necessarily inexact \textbf{due to factors such as asymmetries and unknown physics}.
Computational costs \textbf{make} the training of such models \textbf{challenging; however, advances have been made with attention-based \citep{foundation_attention} and transformer-based \citep{foundation_transformer} models, among others}.
Most prevailing solutions have had to choose either a model driven approach \textbf{(such as this work)}, in which precomputed models are taken as gospel, or a data driven/empirical approach \textbf{(such as Lux \citep{lux})}, in which our knowledge of stellar physics is ignored or treated phenomenologically, depending on the application.
\textbf{Hybrid solutions exist and have been implemented, such as those presented in \citealt{leung2019} and \citealt{rains2024}.
Although a lingering challenge is always how to ideally balance the fusion of model driven and data driven paradigms.}

Spectroscopic surveys such as APOGEE \textbf{\citep{APOGEE}} use their own in-house pipelines to extract stellar parameters from spectra (ASPCAP \textbf{\citep{ASPCAP}} in the case of APOGEE), and these pipelines appear effective.
The core assumption for the vast majority of pipelines is that the data to be analyzed is a list of pixels.
Analysis pipelines may be closed-source or limited to the scope of the survey itself, causing a sort of siloing effect among surveys.
This motivates the development of more universal, instrument-agnostic open-source frameworks that can apply broadly to a range of spectral observations with relatively little tuning.

Multiple efforts have been made in this direction, treating spectra in different ways.
The standard practice is to treat the wavelength and flux as simply two arrays and use bespoke algorithms tailored to a small number of well-calibrated spectral lines to obtain fundamental stellar parameters and chemical compositions \textbf{(e.g. the IGRINS YSO Survey \citep{pipeline1} and the IRTF Spectral Library \citep{pipeline2})}.
Other whole-spectrum fitting abstractions decompose model spectra into an eigenbasis, implementing the data-model comparison stage as a tractable regression, such as \texttt{starfish} \citep{starfish}.

Ideally, we want a system that can self-consistently learn, a genuine AI foundational spectral model.
Such a system would enable the assignment of accurate stellar parameters, and could yield re-usable interpretable spectral models.
\texttt{blase}, first presented in \citealt{blase}, took an important, albeit limited, step in this direction, treating spectra not as a set of pixels or a set of eigenbasis coefficients but as a set of interpretable and traceable spectral lines, specifically Voigt profiles.
Each of these approaches has tradeoffs, but one key scientific advantage of \texttt{blase} comes from its ability to adapt to new information, while preserving some adherence to physics-based models.
This intelligent capability stems from its ability to fit a theoretically unlimited number of nonlinear spectral line parameters with automatic differentiation (autodiff).
Autodiff is a technology that tracks transformations made to data using the chain rule, even being able to differentiate control flow transformations involving if-else blocks, for example.
With the gradient obtained from autodiff's chain rule, we can optimize model parameters by just going against the direction of the gradient (because we usually want to minimize something such as a loss function, we go in the direction of greatest decrease, which is always the negative of the gradient).
Autodiff has been used successfully in other astrophysical contexts such as \texttt{exojax} \textbf{\citep{exojax}} and \texttt{wobble} \citep{wobble}, but the recasting of spectra into sets of inherently nonlinear spectral lines positions \texttt{blase} as a unique and promising semi-empirical tool.
The original \texttt{blase} paper demonstrated the ability to tune spectral lines with autodiff \textbf{and `clone' spectra, recasting them as ML models defined by interpretable sets of tuned spectral lines, however it was} restricted to a pre-selected static synthetic model.

@@

Ideally this process would be monolithic, with the end-to-end spectral inference code powered by a single autodiff\textbf{erenti}able machine learning framework, like PyTorch \textbf{\citep{pytorch}} or JAX \citep{jax}.
However, here we separate the problem into three pieces, only the first \textbf{currently being auto}differentiable.
\textbf{In \autoref{sec:section2},} we scale out the \texttt{blase} method to 1,314 precomputed synthetic spectral model clones, yielding a downloadable archive of pretrained machine learning models with 128,723 unique spectral lines.
\textbf{In \autoref{sec:section3},} we \textbf{then} fit manifolds mapping stellar parameters to uniformly-derived spectral line parameters using regular grid linear interpolation.
Finally, \textbf{in \autoref{sec:section4},} we show how reconstructions of the spectra using said manifolds can be used for inferring stellar fundamental properties from spectra.
This final step resembles the aims of ``atmospheric retrievals'', but in principle should be faster, adaptive, and reusable.
An overview of this process is shown in \autoref{fig:figure1}.

@@

\section{Cloning the PHOENIX Model Grid}
\label{sec:section2}
\subsection{The PHOENIX Subset}
\textbf{For the purposes of this study, w}e chose the widely-adopted PHOENIX synthetic spectral model grid \citep{PHOENIX}.
Our approach can be straightforwardly applied to any other model grid in the future, including substellar atmosphere grids such as Sonora \citep{bobcat, cholla, diamondback, elfowl}, but for now we limit our scope to a \textbf{rectilinear} subset of the PHOENIX grid, focusing on near solar metallicities and a broad range of effective temperature and surface gravity, with details given in \autoref{tab:table1}.
\textbf{This approach is taken due to the computational cost and storage impact of interpolation algorithms.
Future versions with more advanced models will be able to cover a broader scope and bypass the limitation of a rectilinear subset.}

\begin{table*}
    \hspace*{0.6cm}\begin{tabular}{llll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Interval} & \bf{Sampling}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0]& \bf{N/A}\\
        Iron Abundance & [Fe/H] & [-0.5, 0] & \bf{0.5}\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K & \bf{100 K until 7000 K, then 200 K}\\
        Surface Gravity & $\log(g)$ & [2, 6] & \bf{0.5} \\
        Wavelength & $\lambda$ & [8038, 12849] \AA & $\mathbf{R = 500,000}$\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study.
    These limits were imposed to reduce the computational cost of the algorithms and to ensure a rectilinear parameter space in order to work with \texttt{scipy}'s \texttt{RegularGridInterpolator} \citep{scipy}.
    The wavelength limits in particular roughly line up with that of the Habitable Zone Planet Finder (HPF) spectrograph \citep{HPF}.
    This subset is comprised of 1,314 individual spectra: 73 $T_{\mathrm{eff}}$ values, 9 $\log(g)$ values, and 2 [Fe/H] values.}
    \label{tab:table1}
\end{table*}

\subsection{Preprocessing with \texttt{gollum}}
First, the PHOENIX subset was programmatically retrieved with \texttt{gollum} \textbf{\citep{gollum}}, which downloaded \textbf{spectra} from the PHOENIX FTP server. The spectra were then put through a \textbf{short} preprocessing pipeline similar to that from \citealt{blase}.
\begin{enumerate}
    \item \textit{Blackbody Division}: Since the $T_{\mathrm{eff}}$ of each spectrum is known, the \textbf{corresponding} blackbody spectrum was divided out.
    \item \textit{Continuum Normalization}: The spectra were further normalized by dividing them by a \nth{5} order polynomial continuum fit using a peak-finding filter in order to eliminate curvature that would inhibit line modeling.
\end{enumerate}
Mathematically, we can express the preprocessing as follows:
\begin{linenomath*}
\begin{gather}
    \mathsf{\bar{S}} = \dfrac{\mathsf{S}}{\mathsf{B}\mathsf{Q}_5}
\end{gather}
\end{linenomath*}
where $\mathsf{\bar{S}}$ is the preprocessed spectrum, $\mathsf{S}$ is the original spectrum,  $\mathsf{B}$ is the blackbody spectrum, \textbf{and} $\mathsf{Q}_n$ is the $n^\mathrm{th}$ order polynomial continuum fit.
Arithmetic operations between arrays are assumed to be elementwise in all following notation.

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into a physically interpretable intermediate representation: a table of spectral line properties rather than a\textbf{n} array of fluxes.
We used \texttt{blase}, which models spectral lines as Voigt profiles and tunes the profiles to mimic the original PHOENIX spectrum with back propagation.
Back propagation, put simply, is the process of moving in the autodiff gradient field to update ML model parameters (like mentioned earlier, usually against the gradient because we minimize loss functions).
Four parameters were optimized: the line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the Lorentzian width $\gamma$.
The optimization used the Adam optimizer \textbf{\citep{adam}} with a learning rate of 0.05 over 100 epochs.
\textbf{\citealt{blase} recommends this setup as the minimum, and for a proof-of-concept implementation, we found it best to leave it as is.}
In addition, we limited two custom parameters: wing cut to 6,000 and prominence to 0.005.
Wing cut \textbf{(in pixels)} is a parameter that determines the extent of the Voigt profile to evaluate, saving computational resources by not evaluating negligible line wings.
Prominence \textbf{(in normalized flux counts)} sets a lower limit for the amplitude of detected lines, which saves resources by disregarding shallow lines, so in our case we disregard lines \textbf{with amplitude less than} 0.005.
In short, our choices for wing cut and prominence decrease the computational cost of \texttt{blase}'s cloning process at the expense of decreasing its accuracy slightly.
\texttt{blase} uses the pseudo-Voigt approximation, which saves on computational cost while remaining accurate to about 1\% \citep{pseudovoigt}.
The pseudo-Voigt approximation uses a weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voigt profile implementation uses the following:
\begin{linenomath*}
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=0}^2 \mathbf{u}_{n}\left(\dfrac{2\gamma}{f}\right)^{n+1}\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
    \label{eq:equation2}
\end{gather}
\end{subequations}
\end{linenomath*}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and Gaussian profiles, respectively.
Notice that we use $\mu'$ instead of $\mu$ in the formula.
This is because \texttt{blase} optionally allows the line center to shift slightly during optimization, and it is this shifted center which is used in computation.
The individual Voigt profiles are still indexed by $\mu$ for cross-model line identification, explained in the next section.
Once optimization was complete, the list of identified lines was saved to a `state dictionary': a common representation for pre-trained machine learning models that can be stored to disk for reuse later.
These are stored in the \texttt{.pt} file format for each of the 1,314 PHOENIX subset grid points.
The total disk space these files take up is 465 MB (382 MB when downloaded as a zip archive).
For reference, the storage space the PHOENIX subset takes up on disk is approximately 8.1 GB.
This represents a data compression factor of around 20 just by recasting the spectrum with \texttt{blase}.
\textbf{The state dictionaries are available on Zenodo at \url{https://zenodo.org/records/11246174} \citep{zenodo}.}

@@

\textbf{Pre-optimized line centers were required to be equal in order to group lines together.
The values of $\mu$ pre-optimization are recorded by \texttt{blase} to 0.01 \AA \ (picometer) precision; if they instead used the full available floating point precision, this method would likely detect singular (ungroupable) instances of spectral lines across the grid, completely negating the premise of this study.
Of course, this method can be confused by species that produce spectral lines at very close wavelengths and other confounding factors, likely resulting in some false positives and negatives.
However, we deem it sufficient to approximate this way, as an in-depth treatment would be time-consuming and computationally expensive.}

@@

Spectral lines were often only detected in some spectra from the PHOENIX subset.
In \autoref{fig:figure2}, we show that different grid points \textbf{show} differing counts of detected spectral lines.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure2}
    \caption{Number of detected spectral lines at each grid point of the PHOENIX subset.
    We can see that the number of detected lines decreases with increasing $T_{\mathrm{eff}}$.
    \textbf{Remember} that from $T_{\mathrm{eff}} = 7000$ K onward, PHOENIX's sampling increment changes from 100 K to 200 K.
    \textbf{Due to some combination of computational artifacts from our line identification assumptions or innate behavior of PHOENIX grid spectra themselves, outliers such as the obvious one here at $T_{\mathrm{eff}} = 7000$ K and $\log(g) = 6$ dex appear.
    Further exploration into why this happens is not in the scope of this study; for inference we are concerned with the lines themselves, not overall grid point behavior.}}
    \label{fig:figure2}
\end{figure*}

Th\textbf{e number of detected spectral lines changes as a function of stellar parameter} for different reasons.
Perhaps for astrophysical reasons: stellar atmospheres genuinely do not produce that line at detectable strength at the given temperature and surface gravity.
Alternatively, \textbf{\texttt{blase}'s fitting or our line identification assumptions led to artifacts}.
Whatever the cause, these missing lines have immediate practical consequences.
Rectilinear interpolation schemes break in regions where a line does not appear; you can't interpolate a quantity that simply doesn't exist.

To solve this, we artificially populated missing grid points with log-amplitudes of -1000, which retained interpolator stability while nullifying the evaluated line.
\textbf{To prove the feasibility of this computational trick, we can do a simple thought experiment.
Moving 1 K in $T_\text{eff}$ away from a grid point on the boundary of detectability would reduce a line's amplitude by a factor of over 20,000 (or more precisely, $e^{10}$).
This is an extremely large reduction for an extremely small deviation from a grid point, meaning that to any realistic precision, the line immediately vanishes as soon as \texttt{blase} no longer detects it.}

>>>>

\documentclass[twocolumn]{aastex631}
\usepackage{graphicx}
\usepackage{physics}
\usepackage[super]{nth}
\usepackage{enumitem}
\usepackage[all]{hypcap}

@@

\begin{abstract}
    The advent of machine learning (ML) is revolutionary to numerous scientific disciplines, with a growing number of examples in astronomical spectroscopic inference, as ML is more powerful than traditional techniques.
    Here we introduce a hybrid ML (HML) method combining automatic differentiation, interpolation, and Bayesian optimization to infer stellar parameters of stellar spectra.
    We study $T_{\mathrm{eff}}$, $\log(g)$, and [Fe/H], but this method could be extended to other parameters such as [$\alpha$/Fe] (alpha element abundance), C/O (carbon-oxygen ratio), and $f_{sed}$ (sedimentation efficiency).
    We first use \texttt{blase} to semi-empirically recast spectra into sets of Voigt profiles.
    \texttt{blase} is run on 1,314 synthetic spectra from a rectilinear subset of the PHOENIX model grid ($T_\text{eff}$: [2300, 12000] K, $\log(g)$: [2, 6], [Fe/H]: [-0.5, 0], $\lambda$: [8038, 12849] \AA).
    For 128,723 detected features, we map stellar parameters to spectral line parameters using linear interpolation.
    This creates the PHOENIX generator, enabling parallelized spectral synthesis.
    Gaussian Process minimization is used to infer stellar parameters by minimizing a root-mean-square loss function.
    Testing 210 noise-free models ($T_{\mathrm{eff}}$: [3000, 11000] K, $\log(g)$: [2, 6], [Fe/H]: [-0.5, 0]), we find inference errors: $T_{\mathrm{eff}}$: 93 K, $\log(g)$: 0.24, and [Fe/H]: 0.056 for $T_\text{eff} < $ 7000 K, and $T_{\mathrm{eff}}$: 347 K, $\log(g)$: 0.26, and [Fe/H]: 0.16 for $T_\text{eff} \ge$ 7000 K.
    We also upload online an archive of \texttt{blase} models of the PHOENIX subset.
    This proof-of-concept study shows that semi-empirical HML is a viable alternative to traditional approaches in spectroscopic inference.
\end{abstract}

@@

Modern astronomical spectroscopy takes advantage of intuitive, performant spectral models such as KORG \citep{korg}, however it would further benefit from the paradigm of interpretability.
The determination of fundamental properties, the creation of Extreme Precision Radial Velocity (EPRV) templates, and the application of composite spectral fitting could all leverage such an innovation.
This ambitious aim may be broadly referred to as a ``foundational spectroscopy model for astrophysics'', in reference to the same category of models used for large language models (LLMs) in artificial intelligence (AI).
There are many challenges with creating such a foundational spectral model for astrophysics.
Physical inputs into the spectral modeling process are imperfect, and simplifying assumptions in stellar (and substellar) atmospheres are necessarily inexact due to factors such as asymmetries and unknown physics.
Computational costs make the training of such models challenging; however, advances have been made with attention-based \citep{foundation_attention} and transformer-based \citep{foundation_transformer} models, among others.
Most prevailing solutions have had to choose either a model driven approach (such as this work), in which precomputed models are taken as gospel, or a data driven/empirical approach (such as Lux \citep{lux}), in which our knowledge of stellar physics is ignored or treated phenomenologically, depending on the application.
Hybrid solutions exist and have been implemented, such as those presented in \citealt{leung2019} and \citealt{rains2024}.
Although a lingering challenge is always how to ideally balance the fusion of model driven and data driven paradigms.

Spectroscopic surveys such as APOGEE \citep{APOGEE} use their own in-house pipelines to extract stellar parameters from spectra (ASPCAP \citep{ASPCAP} in the case of APOGEE), and these pipelines appear effective.
The core assumption for the vast majority of pipelines is that the data to be analyzed is a list of pixels.
Analysis pipelines may be closed-source or limited to the scope of the survey itself, causing a sort of siloing effect among surveys.
This motivates the development of more universal, instrument-agnostic open-source frameworks that can apply broadly to a range of spectral observations with relatively little tuning.

Multiple efforts have been made in this direction, treating spectra in different ways.
The standard practice is to treat the wavelength and flux as simply two arrays and use bespoke algorithms tailored to a small number of well-calibrated spectral lines to obtain fundamental stellar parameters and chemical compositions (e.g. the IGRINS YSO Survey \citep{pipeline1} and the IRTF Spectral Library \citep{pipeline2}).
Other whole-spectrum fitting abstractions decompose model spectra into an eigenbasis, implementing the data-model comparison stage as a tractable regression, such as \texttt{starfish} \citep{starfish}.

Ideally, we want a system that can self-consistently learn, a genuine AI foundational spectral model.
Such a system would enable the assignment of accurate stellar parameters, and could yield re-usable interpretable spectral models.
\texttt{blase}, first presented in \citealt{blase}, took an important, albeit limited, step in this direction, treating spectra not as a set of pixels or a set of eigenbasis coefficients but as a set of interpretable and traceable spectral lines, specifically Voigt profiles.
Each of these approaches has tradeoffs, but one key scientific advantage of \texttt{blase} comes from its ability to adapt to new information, while preserving some adherence to physics-based models.
This intelligent capability stems from its ability to fit a theoretically unlimited number of nonlinear spectral line parameters with automatic differentiation (autodiff).
Autodiff is a technology that tracks transformations made to data using the chain rule, even being able to differentiate control flow transformations involving if-else blocks, for example.
With the gradient obtained from autodiff's chain rule, we can optimize model parameters by just going against the direction of the gradient (because we usually want to minimize something such as a loss function, we go in the direction of greatest decrease, which is always the negative of the gradient).
Autodiff has been used successfully in other astrophysical contexts such as \texttt{exojax} \citep{exojax} and \texttt{wobble} \citep{wobble}, but the recasting of spectra into sets of inherently nonlinear spectral lines positions \texttt{blase} as a unique and promising semi-empirical tool.
The original \texttt{blase} paper demonstrated the ability to tune spectral lines with autodiff and `clone' spectra, recasting them as ML models defined by interpretable sets of tuned spectral lines, however it was restricted to a pre-selected static synthetic model.

@@

Ideally this process would be monolithic, with the end-to-end spectral inference code powered by a single autodifferentiable machine learning framework, like PyTorch \citep{pytorch} or JAX \citep{jax}.
However, here we separate the problem into three pieces, only the first currently being autodifferentiable.
In \autoref{sec:section2}, we scale out the \texttt{blase} method to 1,314 precomputed synthetic spectral model clones, yielding a downloadable archive of pretrained machine learning models with 128,723 unique spectral lines.
In \autoref{sec:section3}, we then fit manifolds mapping stellar parameters to uniformly-derived spectral line parameters using regular grid linear interpolation.
Finally, in \autoref{sec:section4}, we show how reconstructions of the spectra using said manifolds can be used for inferring stellar fundamental properties from spectra.
This final step resembles the aims of ``atmospheric retrievals'', but in principle should be faster, adaptive, and reusable.
An overview of this process is shown in \autoref{fig:figure1}.

@@

\section{Cloning the PHOENIX Model Grid}
\label{sec:section2}
\subsection{The PHOENIX Subset}
For the purposes of this study, we chose the widely-adopted PHOENIX synthetic spectral model grid \citep{PHOENIX}.
Our approach can be straightforwardly applied to any other model grid in the future, including substellar atmosphere grids such as Sonora \citep{bobcat, cholla, diamondback, elfowl}, but for now we limit our scope to a rectilinear subset of the PHOENIX grid, focusing on near solar metallicities and a broad range of effective temperature and surface gravity, with details given in \autoref{tab:table1}.
This approach is taken due to the computational cost and storage impact of interpolation algorithms.
Future versions with more advanced models will be able to cover a broader scope and bypass the limitation of a rectilinear subset.

\begin{table*}
    \hspace*{1cm}\begin{tabular}{llll}
        \hline
        \bf{Parameter} & \bf{Symbol} & \bf{Interval} & \bf{Sampling}\\
        \hline
        Alpha Element Abundance & $\alpha$ & [0]& N/A\\
        Iron Abundance & [Fe/H] & [-0.5, 0] & 0.5\\
        Effective Temperature & $T_{\mathrm{eff}}$ & [2300, 12000] K & 100 K until 7000 K, then 200 K\\
        Surface Gravity & $\log(g)$ & [2, 6] & 0.5 \\
        Wavelength & $\lambda$ & [8038, 12849] \AA & $R = 500,000$\\
        \hline
    \end{tabular}
    \caption{The subset of the PHOENIX grid used in this study.
    These limits were imposed to reduce the computational cost of the algorithms and to ensure a rectilinear parameter space in order to work with \texttt{scipy}'s \texttt{RegularGridInterpolator} \citep{scipy}.
    The wavelength limits in particular roughly line up with that of the Habitable Zone Planet Finder (HPF) spectrograph \citep{HPF}.
    This subset is comprised of 1,314 individual spectra: 73 $T_{\mathrm{eff}}$ values, 9 $\log(g)$ values, and 2 [Fe/H] values.}
    \label{tab:table1}
\end{table*}

\subsection{Preprocessing with \texttt{gollum}}
First, the PHOENIX subset was programmatically retrieved with \texttt{gollum} \citep{gollum}, which downloaded spectra from the PHOENIX FTP server. The spectra were then put through a short preprocessing pipeline similar to that from \citealt{blase}.
\begin{enumerate}
    \item \textit{Blackbody Division}: Since the $T_{\mathrm{eff}}$ of each spectrum is known, the corresponding blackbody spectrum was divided out.
    \item \textit{Continuum Normalization}: The spectra were further normalized by dividing them by a \nth{5} order polynomial continuum fit using a peak-finding filter in order to eliminate curvature that would inhibit line modeling.
\end{enumerate}
Mathematically, we can express the preprocessing as follows:
\begin{linenomath*}
\begin{gather}
    \mathsf{\bar{S}} = \dfrac{\mathsf{S}}{\mathsf{B}\mathsf{Q}_5}
\end{gather}
\end{linenomath*}
where $\mathsf{\bar{S}}$ is the preprocessed spectrum, $\mathsf{S}$ is the original spectrum,  $\mathsf{B}$ is the blackbody spectrum, and $\mathsf{Q}_n$ is the $n^\mathrm{th}$ order polynomial continuum fit.
Arithmetic operations between arrays are assumed to be elementwise in all following notation.

\subsection{Line Identification with \texttt{blase}}
The next step was to convert the PHOENIX subset into a physically interpretable intermediate representation: a table of spectral line properties rather than an array of fluxes.
We used \texttt{blase}, which models spectral lines as Voigt profiles and tunes the profiles to mimic the original PHOENIX spectrum with back propagation.
Back propagation, put simply, is the process of moving in the autodiff gradient field to update ML model parameters (like mentioned earlier, usually against the gradient because we minimize loss functions).
Four parameters were optimized: the line center $\mu$, the log-amplitude $\ln(a)$, the Gaussian width $\sigma$, and the Lorentzian width $\gamma$.
The optimization used the Adam optimizer \citep{adam} with a learning rate of 0.05 over 100 epochs.
\citealt{blase} recommends this setup as the minimum, and for a proof-of-concept implementation, we found it best to leave it as is.
In addition, we limited two custom parameters: wing cut to 6,000 and prominence to 0.005.
Wing cut (in pixels) is a parameter that determines the extent of the Voigt profile to evaluate, saving computational resources by not evaluating negligible line wings.
Prominence (in normalized flux counts) sets a lower limit for the amplitude of detected lines, which saves resources by disregarding shallow lines, so in our case we disregard lines with amplitude less than 0.005.
In short, our choices for wing cut and prominence decrease the computational cost of \texttt{blase}'s cloning process at the expense of decreasing its accuracy slightly.
\texttt{blase} uses the pseudo-Voigt approximation, which saves on computational cost while remaining accurate to about 1\% \citep{pseudovoigt}.
The pseudo-Voigt approximation uses a weighted average of a Gaussian and Lorentzian as opposed to a convolution.
\texttt{blase}'s pseudo-Voigt profile implementation uses the following:
\begin{linenomath*}
\begin{subequations}
\begin{gather}
    \tilde{\mathsf{V}}_\mu(\lambda) = a\left[\eta \mathbf{L}(\lambda - \mu'; f) + (1 - \eta)\mathbf{G}(\lambda - \mu'; f)\right]\\
    \eta = \sum_{n=0}^2 \mathbf{u}_{n}\left(\dfrac{2\gamma}{f}\right)^{n+1}\\
    f = 32\sum_{n=0}^5 \mathbf{v}_n \left(\sqrt{2\ln(2)}\sigma\right)^{5-n}(\gamma)^n\\
    \nonumber \mathbf{u} = \mqty[1.36603 \\ -0.47719 \\ 0.11116] \quad \mathbf{v} = \mqty[1 \\ 2.69269 \\ 2.42843 \\ 4.47163 \\ 0.07842 \\ 1]
    \label{eq:equation2}
\end{gather}
\end{subequations}
\end{linenomath*}
where $\mathbf{L}$ and $\mathbf{G}$ are abbreviations for Lorentzian and Gaussian profiles, respectively.
Notice that we use $\mu'$ instead of $\mu$ in the formula.
This is because \texttt{blase} optionally allows the line center to shift slightly during optimization, and it is this shifted center which is used in computation.
The individual Voigt profiles are still indexed by $\mu$ for cross-model line identification, explained in the next section.
Once optimization was complete, the list of identified lines was saved to a `state dictionary': a common representation for pre-trained machine learning models that can be stored to disk for reuse later.
These are stored in the \texttt{.pt} file format for each of the 1,314 PHOENIX subset grid points.
The total disk space these files take up is 465 MB (382 MB when downloaded as a zip archive).
For reference, the storage space the PHOENIX subset takes up on disk is approximately 8.1 GB.
This represents a data compression factor of around 20 just by recasting the spectrum with \texttt{blase}.
The state dictionaries are available on Zenodo at \url{https://zenodo.org/records/11246174} \citep{zenodo}.

@@

Pre-optimized line centers were required to be equal in order to group lines together.
The values of $\mu$ pre-optimization are recorded by \texttt{blase} to 0.01 \AA \ (picometer) precision; if they instead used the full available floating point precision, this method would likely detect singular (ungroupable) instances of spectral lines across the grid, completely negating the premise of this study.
Of course, this method can be confused by species that produce spectral lines at very close wavelengths and other confounding factors, likely resulting in some false positives and negatives.
However, we deem it sufficient to approximate this way, as an in-depth treatment would be time-consuming and computationally expensive.

@@

Spectral lines were often only detected in some spectra from the PHOENIX subset.
In \autoref{fig:figure2}, we show that different grid points show differing counts of detected spectral lines.

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure2}
    \caption{Number of detected spectral lines at each grid point of the PHOENIX subset.
    We can see that the number of detected lines decreases with increasing $T_{\mathrm{eff}}$.
    Remember that from $T_{\mathrm{eff}} = 7000$ K onward, PHOENIX's sampling increment changes from 100 K to 200 K.
    Due to some combination of computational artifacts from our line identification assumptions or innate behavior of PHOENIX grid spectra themselves, outliers such as the obvious one here at $T_{\mathrm{eff}} = 7000$ K and $\log(g) = 6$ dex appear.
    Further exploration into why this happens is not in the scope of this study; for inference we are concerned with the lines themselves, not overall grid point behavior.}
    \label{fig:figure2}
\end{figure*}

The number of detected spectral lines changes as a function of stellar parameter for different reasons.
Perhaps for astrophysical reasons: stellar atmospheres genuinely do not produce that line at detectable strength at the given temperature and surface gravity.
Alternatively, \texttt{blase}'s fitting or our line identification assumptions led to artifacts.
Whatever the cause, these missing lines have immediate practical consequences.
Rectilinear interpolation schemes break in regions where a line does not appear; you can't interpolate a quantity that simply doesn't exist.

To solve this, we artificially populated missing grid points with log-amplitudes of -1000, which retained interpolator stability while nullifying the evaluated line.
To prove the feasibility of this computational trick, we can do a simple thought experiment.
Moving 1 K in $T_\text{eff}$ away from a grid point on the boundary of detectability would reduce a line's amplitude by a factor of over 20,000 (or more precisely, $e^{10}$).
This is an extremely large reduction for an extremely small deviation from a grid point, meaning that to any realistic precision, the line immediately vanishes as soon as \texttt{blase} no longer detects it.

Examples of the appearance of missing sections in heatmaps where a line does not appear can be seen in \autoref{fig:figure3} and \autoref{fig:figure4}, which show heatmaps of the 11617.66 \AA \ C I and the 10123.66 \AA \ He II lines (identified with NIST \citep{NIST}) varying over a solar metallicity slice of the PHOENIX subset stellar parameter space.
Across the entire PHOENIX subset, \texttt{blase} detected 128,723 spectral lines with unique $\mu$.
On average, a PHOENIX spectrum has 9,167 detected spectral lines, with the minimum being 252 ($T_\text{eff}$ = 12000 K, $\log(g)$ = 6, [Fe/H] = -0.5) and the maximum being 34,551 ($T_\text{eff}$ = 2300 K, $\log(g)$ = 3, [Fe/H] = 0).
Every one of these lines can be visualized as a manifold mapping a 3D stellar parameter vector to a 4D spectral line parameter vector, and every one was interpolated to map stellar parameters to spectral line properties.
\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figure3}
    \caption{Heatmaps showing how $\ln(a)$, $\sigma$, $\gamma$, and $\mu' - \mu$ vary over the PHOENIX subset slice at solar metallicity of a C I spectral line, notable for being detected at nearly all subset grid points.
    Notice the missing chunk in the top left; this could perhaps be due to the line center shifting to the point that \texttt{blase} identified the chunk as a different line entirely (notice the line center offset increasing in magnitude approaching the chunk border).
    Since \texttt{blase} did not detect a spectral line here, we artificially populate those points with lines that have $\ln(a) = -1000$ during interpolation.}
    \label{fig:figure3}
\end{figure*}
\begin{figure}
    \centering
    \includegraphics[width=\columnwidth]{figure4}
    \caption{Heatmaps showing how $\ln(a)$, $\sigma$, $\gamma$, and $\mu' - \mu$ vary over the PHOENIX subset slice at solar metallicity of a He II spectral line.
    \texttt{blase} detects this line at a select chunk of grid points, leading to data only existing between $T_\text{eff}$ 3600 K and 6600 K}.
    \label{fig:figure4}
\end{figure}
